---
title: "P8108 Group 2 Survival Analysis Project"
author: 
  - Yiming Zhao (yz3955) 
  - Wenshan Qu (wq2160)
  - Tucker Morgan (tlm2152)
  - Junzhe Shao (js5959)
  - Benjamin Goebel (bpg2118)
date: "2022-10-19"
output: pdf_document
---

```{r message=FALSE}
library(survival)
library(tidyverse)
library(tidymodels)
```

## Train Test Split
```{r}
set.seed(2022)

rotterdam_split <- initial_split(rotterdam, prop = 0.8, strata = death)
rotterdam_training <- training(rotterdam_split)
rotterdam_test <- testing(rotterdam_split)
```

## Perform 10-fold Cross-Validation
```{r}
set.seed(2022)

rotterdam_folds <- vfold_cv(rotterdam_training, v = 10, repeats = 5, 
                            strata = death)

rotterdam_folds <- rotterdam_folds %>%
  mutate(split_analysis = map(splits, analysis),
         split_assessment = map(splits, assessment))
```

## Introduction

## Methods

## Exploratory Data Analysis

## Cross-Validation

## Cox/Cox with elastic net

## Random survival forest

## Conformalized survival analysis

## Supplemental analyses

## Results

## Exploratory Data Analysis

## Cross Validation

## Cox/Cox with elastic net

## Random survival forest

## Conformalized survival analysis

## Supplemental analyses

## Discussion

## How our results compare with past research

## Conclusion